---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# tigris-basemap

<!-- badges: start -->

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental) [![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT) [![Project Status: Concept – Minimal or no implementation has been done yet, or the repository is only intended to be a limited example, demo, or proof-of-concept.](https://www.repostatus.org/badges/latest/concept.svg)](https://www.repostatus.org/#concept)

<!-- badges: end -->

The goal of tigris-basemap is to share a reproducible data pipeline (RAP) built with `{targets}` and `{tigris}` for creating static basemaps of counties and metropolitan areas in the United States of America using spatial data from the Census Bureau and USGS.

At present this pipeline relies on development versions of `{tigris}` and `{rmapshaper}`. These can be installed using `{pak}` or `{devtools}`:

```{r eval=FALSE}
pak::pkg_install("elipousson/tigris")
pak::pkg_install("elipousson/rmapshaper")
```

## Configuring geographies

The geographies used by the pipeline are specified in `geographies.yml`. Each entry under the `geographies` key creates a county basemap and a regional basemap for a focal county and the metropolitan statistical area (MSA) that contains it:

```yaml
geographies:
  baltimore_city:
    name: Baltimore City
    description: County and regional basemaps for Baltimore City, Maryland.
    tigris_year: 2023
    state: MD
    county: Baltimore city
    msa: Baltimore-Columbia-Towson, MD Metro Area
    division:
      type: custom
      url: https://services1.arcgis.com/mVFRs7NF4iFitgbY/arcgis/rest/services/Community_Statistical_Areas_(CSAs)__Reference_Boundaries/FeatureServer/0
      name_col: Community
      snap_to_tracts: true
```

- `name`, `tigris_year`, `state`, `county`, and `msa` are required. `msa` must use the name of the MSA from the American Community Survey (ACS) and `state` is the state of the focal county.
- MSAs can include multiple states, e.g. `Washington-Arlington-Alexandria, DC-VA-MD-WV Metro Area`. The states in the MSA are parsed from the name or can be listed with the optional `msa_states` field (which must match the name). The counties in the MSA are selected using the CBSA code (`CBSAFP`) for each county, and the pipeline checks that the focal county is in the MSA and that the selected counties cover the MSA.
- `division` is optional. `type` can be `tract` or `zcta` (both loaded with `{tigris}`) or `custom`. Custom divisions require either a `path` to a file readable by `sf::read_sf()` or a `url` for an ArcGIS layer. Set `snap_to_tracts: true` with a `name_col` to rebuild custom divisions from census tracts.

Target names use the entry key as a suffix, e.g. `county_basemap_baltimore_city`. Data shared by multiple geographies is only loaded once: national data for each year (e.g. `us_msa_2023`), state data for each state and year, including every state in each MSA (e.g. `counties_md_2023` is used for both the Baltimore and Washington, DC MSAs), and MSA data for each MSA and year (e.g. `msa_water_baltimore_md_2023`). To build the targets for a single geography, use:

```{r eval=FALSE}
targets::tar_make(names = tidyselect::ends_with("_baltimore_city"))
```

## Configuring data sources and basemap styles

Every external data source used by the pipeline is listed in `sources.yml` with a description, the `{tigris}` function or ArcGIS layer URL used to load it, and optional `args` for the loader function and default `processing_steps`:

```yaml
sources:
  usgs_pad:
    name: USGS Protected Areas Database (PAD-US) by manager name
    description: >-
      Protected areas loaded with load_arc_url() and filtered by the area of
      interest. Small parcels are dropped on the server with a `where` clause
      on GIS_Acres.
    type: arcgis
    url: https://services.arcgis.com/v01gqwM5QqNysAAi/ArcGIS/rest/services/Manager_Name/FeatureServer/0
    args:
      where: GIS_Acres >= 20
    processing_steps:
      - name: st_make_valid
```

Source `processing_steps` (such as repairing invalid geometry) always run first. Loaders only download data and transform it to the basemap CRS: clipping and other changes to geometry are explicit processing steps.

The data, view, and style layers for the county and MSA basemaps are specified in `basemaps.yml`. Style layers follow the [MapLibre Style Spec](https://maplibre.org/maplibre-style-spec/layers/) and are drawn in order with `ggplot2::geom_sf()`. Each layer can use `processing_steps` named after the function they call (e.g. `ms_simplify` or `st_buffer`) to prepare geometry for cartographic purposes:

```yaml
basemaps:
  msa:
    data:
      water:
        source: tigris_area_water
        processing_steps:
          - name: ms_clip
            args:
              clip:
                source: msa_counties
              remove_slivers: true
    staging:
      metro_areas:
        source: tigris_metro_divisions
    view:
      bounds:
        source: [counties, buffer, water]
    layers:
      - id: water
        type: fill
        source: water
        processing_steps:
          - name: ms_simplify
            args:
              keep: 0.02
          - name: st_union
        paint:
          fill-color: "#66B7C9"
```

Data that is loaded by the pipeline but not yet used by a basemap is listed under `staging`. See [`docs/style-spec.md`](docs/style-spec.md) for the full schema, the supported processing steps, and how MapLibre properties map onto ggplot2 arguments.

Settings from both files are inserted directly into the target commands, so changing the settings for one source or layer only invalidates the targets that use it. Changes to `paint` or `layout` properties only invalidate the basemap targets (e.g. `msa_basemap_baltimore_city`) without re-running any processing steps.

## Examples

Here is an example of a county map created with this pipeline:

```{r}
targets::tar_read(county_basemap_baltimore_city)
```

This map does not include the outside border of the county as it is intended to be used in combination with a data layer and then county boundary applied as a foreground layer.

Here is an example of a regional map created with this pipeline:

```{r}
targets::tar_read(msa_basemap_baltimore_city)
```

The design of both maps are created for use in Baltimore City, Maryland and the Baltimore-Columbia-Towson, MD Metro Area. The pipeline works with other counties and metro areas but has not been tested extensively and likely will not work with all geographies.
